Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field

### 2024-04-28 paper/ms.tex
update ms

% <250 words
\begin{abstract}
  Recent studies of stars in open clusters have shown how the rotation
  rates and lithium content of FGK stars evolve during the first few
  billion years of their lives.
  %
  Here we apply this knowledge to stars observed by the
  primary Kepler mission, and calculate the implied rotation and
  lithium-based ages.
  %
  Rotation periods are drawn from previous Kepler analyses; lithium is
  measured from new and archival HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which we believe
  gyrochronology is valid.
  %
  The independent lithium ages show X,Y,Z, though they are subject to
  their own systematic issues.
  %
  The relative statistical precisions from rotation, $\sigma_{t_{\rm
  gyro}} / t_{\rm gyro}$, are XX +YY -ZZ\%, which is X.X +Y.Y -Z.Z more
  statistically precise than isochronal ages for the same stars, and
  $\approx$X.X times less precise than cluster-based ages.
  %
  The results include \nplyounggyrotwosigma\ planets younger than
  1\,Gyr at 2$\sigma$ (\nplyounggyrothreesigma\ planets at 3$\sigma$).
  %
  While lithium provides only marginal added information for most of
  this sample, we report new robust ages from both rotation and
  lithium for 14 confirmed $<$500\,Myr planets, including five
  mini-Neptunes, five super-Earths, three Earths and sub-Earths, and a
  Neptune-sized planet.
  %
  In the broader Kepler target star sample, we find that the
  gyrochrone age distribution rises linearly from 0 to 3\,Gyr, with
  stars 2\,Gyr old being twice as common in the Kepler field as stars
  1\,Gyr old.
  %
  This parent stellar age distribution bounds the relative balance of
  young and old planets detectable not only by Kepler, but also by
  future wide-field transit surveys.
  %
\end{abstract}

@@

\subsection{Age distribution}

@@

@@

\startlongtable
\begin{deluxetable*}{lllllllrrrrrrl}
	%\tabletypesize{\footnotesize}
	\tabletypesize{\scriptsize}
  \tablecaption{Ages of Kepler planets and planet candidates.  This
  version of the table is truncated to include the youngest fifty
  confirmed Kepler planets, and the twenty youngest candidates.  The
  full machine-readable table contains ages and age limits for
  \nnonfopkoissomeageinfo\ non-false positive KOIs with MES$>$10.  
  \label{tab:planets}}
	%\toprule
	%\midrule
	%\endhead
	\startdata
  KOI & Kepler &  $T_{\rm eff}$ & $P_{\rm rot}$ & EW$_{\rm Li}$ &
  $t_{\rm gyro}$ & $t_{\rm Li} $ & $R_{\rm p}$ & $P_{\rm orb}$ &
  $Q_{\rm planet}$ & $Q_{\rm gyro}$ &  Spec? & Comment \\
  -- &   -- & K & days & m\AA & Myr & Myr &    Earths &    days &       int  & int & bool & -- \\
	\hline
	\input{table_subgyr_confirmed} %\\
%	\rule
	\input{table_subgyr_candidate}
	\enddata
	\tablecomments{Rows
		are sorted by the minimum of either the rotation or lithium-based
		age, and ``confirmed'' planets appear before``candidate'' planets.
		The bit quality flags for the rotation-based ages,
		$Q_{\rm gyro}$, are described in Section~\ref{subsec:flags}.
		Concisely summarized, they are:
		Bit 0: $T_{\rm eff}/{\rm K} \in [ 3800-6200]$ ?
		Bit 1: $\log g$$<$4.2?
		Bit 2: $M_{\rm G}$<3.9?
		Bit 3: In KEBC?
		Bit 4: Large $d_{\rm xm,Kep-Gaia}$?
		Bit 5: Confused Kep-Gaia crossmatch?
		Bit 6: Gaia DR3 non-single star?
		Bit 7: RUWE$>$1.4?
		Bit 8: Crowded?
		Bit 9: Far from main-sequence?
		Bit 10: \citetalias{Santos_2021} CP/CB?
		As an example, Kepler-1627 has $Q_{\rm gyro}$ flagged with bit 10 and bit 7.
		The analogous planet quality bit-flag, $Q_{\rm planet}$, has the
		following meaning.
		Bit 0: Candidate not reliable?
		Bit 1: MES$<$10?
		Bit 2: Grazing?
	}
\end{deluxetable*}

>>>>

% <250 words
\begin{abstract}
  Recent studies of stars in open clusters have shown how the rotation
  rates and lithium content of FGK stars evolve during the first few
  billion years of their lives.
  %
  Here we apply this knowledge to stars observed by the primary Kepler
  mission to calculate rotation and lithium-based ages.
  %
  Rotation periods are drawn from previous Kepler analyses; lithium is
  measured from new and archival HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which we believe
  gyrochronology is valid;
  these rotation-based ages are $\approx$15-20$\times$ more precise
  than isochrone-based ages for our sample of FGK stars.
  %
  The independent lithium ages are consistent for XX\% of cases;
  however this method is less precise than rotation-based age dating
  for most of our sample.
  %
  Our resulting catalog of planetary ages includes
  \nplyounggyrotwosigma\ planets younger than 1\,Gyr at 2$\sigma$
  (\nplyounggyrothreesigma\ planets at 3$\sigma$).
  %
  In the broader sample of stars in the Kepler field, we find that the
  rotation-based age distribution decreases linearly from 3\,Gyr ago
  to present day, implying a linearly decreasing star formation rate.
  %
  While this parent stellar age distribution has implications for the
  star formation history of our Galaxy, it also bounds the relative
  balance of young and old planets detectable using both Kepler and
  also future wide-field transit surveys.
  %
\end{abstract}

@@

\subsection{The Demographic Cliff}

@@

While lithium provides only marginal added information for most of
this sample, we report new robust ages from both rotation and
lithium for 14 confirmed $<$500\,Myr planets, including five
mini-Neptunes, five super-Earths, three Earths and sub-Earths, and a
Neptune-sized planet.

@@

\startlongtable
\begin{deluxetable*}{llllllllrrrrrrl}
	%\tabletypesize{\footnotesize}
	\tabletypesize{\scriptsize}
  \tablecaption{Ages of Kepler planets and planet candidates.  This
  version of the table is truncated to include the youngest fifty
  confirmed Kepler planets, and the twenty youngest candidates,
  sorted by the minimum of either the rotation or lithium-based
  age.  The
  full machine-readable table contains ages and age limits for
  \nnonfopkoissomeageinfo\ non-false positive KOIs with MES$>$10.  
  \label{tab:planets}}
	%\toprule
	%\midrule
	%\endhead
	\startdata
  KOI & Kepler &  $T_{\rm eff}$ & $P_{\rm rot}$ & EW$_{\rm Li}$ &
  $t_{\rm gyro}$ & $t_{\rm Li} $ & Consistent? & $R_{\rm p}$ & $P_{\rm orb}$ &
  $Q_{\rm planet}$ & $Q_{\rm gyro}$ &  Spec? & Comment \\
  -- &   -- & K & days & m\AA & Myr & Myr &  str &   Earths &    days &       int  & int & bool & -- \\
	\hline
	\input{table_subgyr_confirmed} %\\
%	\rule
	\input{table_subgyr_candidate}
	\enddata
	\tablecomments{``Confirmed'' planets appear before``candidate'' planets.
		The bit quality flags for the rotation-based ages,
		$Q_{\rm gyro}$, are described in Section~\ref{subsec:flags}.
		Concisely summarized, they are:
		Bit 0: $T_{\rm eff}/{\rm K} \in [ 3800-6200]$ ?
		Bit 1: $\log g$$<$4.2?
		Bit 2: $M_{\rm G}$<3.9?
		Bit 3: In KEBC?
		Bit 4: Large $d_{\rm xm,Kep-Gaia}$?
		Bit 5: Confused Kep-Gaia crossmatch?
		Bit 6: Gaia DR3 non-single star?
		Bit 7: RUWE$>$1.4?
		Bit 8: Crowded?
		Bit 9: Far from main-sequence?
		Bit 10: \citetalias{Santos_2021} CP/CB?
		As an example, Kepler-1627 has $Q_{\rm gyro}$ flagged with bit 10 and bit 7.
		The analogous planet quality bit-flag, $Q_{\rm planet}$, has the
		following meaning.
		Bit 0: Candidate not reliable?
		Bit 1: MES$<$10?
		Bit 2: Grazing?
	}
\end{deluxetable*}
